\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/cv.tex
Comments Marc
Comments François, smearing figure
Compiling
Update bttda formulation
cover page
Proofread

\newenvironment{cvlist}{
	\begin{description}[font=\normalfont]%\setlength\itemsep{0em}
}{
  \end{description}
}

@@

% Curriculum vitae =============================================================
\chapter{Curriculum vit\ae}

\subsection*{About me}
I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

@@

\begin{cvlist}
	\cvitem{Ph.D.\ in Biomedical Sciences}{KU Leuven, 2024}
	\cvitem{Ph.D.\ in Control Science and Signal Processing}{University of
		Lille, 2024}
	\cvitem{M.Sc.\ in Engineering Science: Computer Science}{KU Leuven, 2024}
	\cvitem{B.Sc.\ in Informatics}{KU Leuven, 2017}
\end{cvlist}

\subsection*{Publications}
\nocite{%
	VanDenKerchove2020,
	Libert2022,
	VanDenKerchove2022,
	VanDenKerchove2024,
}
\printbibliography[heading=none, resetnumbers=true]

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
\end{cvlist}

>>>>

\newenvironment{cvlist}{%
	\begin{description}[font=\normalfont]%\setlength\itemsep{0em}
}{%
  \end{description}
}

@@

% Curriculum vitae =============================================================
\chapter*{Curriculum vit\ae}
Arne Van Den Kerchove \\
\texttt{arne@vandenkerchove.com}

\subsection*{About me}
I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

@@

\begin{cvlist}
	\cvitem{PhD\ in Biomedical Sciences}{KU Leuven, 2024}
	\cvitem{PhD\ in Control Science and Signal Processing}{University of
		Lille, 2024}
	\cvitem{M.Sc.\ in Engineering Science: Computer Science}{KU Leuven, 2024}
	\cvitem{B.Sc.\ in Informatics}{KU Leuven, 2017}
\end{cvlist}

\subsection*{Publications}
\begin{refsection}
\setmaxbibnames{9999}
\nocite{%
	VanDenKerchove2020,
	Libert2022,
	VanDenKerchove2022,
	VanDenKerchove2024,
	VanDenKerchove2024a,
}
\printbibliography[heading=none, resetnumbers=true]
\end{refsection}

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global PhD Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
\end{cvlist}

### 2024-10-08 chapters/funding.tex
Proofread

\chapter*{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}

@@

We also thank,  Arno Libert,
Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
for the works presented in \autoref{sec:covert-align} and
\autoref{sec:patients}.
and in general the colleagues at the KU Leuven Computational Neurosciences
group at KU Leuven and the BCI team at the University of Lille.
Furthermore, we thank the support teams at TRAINM Neuro Rehab Clinics Antwerp,
the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in~\autoref{sec:bttda}.

\todo{people at TRAINM, MAS and Kristl Claeys}
\todo{juliette meunier}

>>>>

\chapter*{Funding \& acknowledgments}

@@

We also thank  Arno Libert,
Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
for the works presented in \autoref{sec:covert-align} and
\autoref{sec:patients}.
and in general the colleagues at the KU Leuven Computational Neurosciences
group at KU Leuven and the BCI team at the University of Lille.
Furthermore, we thank the support teams at TRAINM Neuro Rehab Clinics Antwerp,
the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in~\autoref{sec:bttda}.

### 2024-10-08 chapters/gaze_independence.tex
Proofread

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer
to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus,
\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).},
other eye tremors
\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.},
gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.},
involuntary movements, impaired pursuit, \ldots), and even complete
ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only affects vision and coordination but also
their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  \input{chapters/gaze_independence_incidence.tex}
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

@@

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or it is very limited,
    such as the ability to communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

The most severely impaired of these patients constitute the \ac{lis}
patient group, which forms one of the main \ac{bci} interest groups.
In this work, we use the term \ac{lis} for a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
\todo{[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
visual function of patients with locked-in syndrome: report
of 13 cases. J Fr Ophtalmol. 2016;39:437–440.}
Hence, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.

@@

The general goal of this PhD is to tackle the gaze-dependence problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from gaze position, which do not necessarily need to coincide.

@@

To address this, we improved upon an in-house developed, state-of-the-art \ac{erp}
decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(STBF-struct).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, yielding insights for our next
efforts to cope with the non-stationarity of the P3 signal component.
This approach has been published by~\textcite{VanDenKerchove2022} and will be
described in \autoref{sec:stbf-struct}.

@@

While \ac{hoda} is effective, it is limited by its reliance on a rigid core
structure, which may not fully capture the complexity of neural data in some
scenarios. To address this, \ac{bttda}
introduces a more flexible block-term structure.
\ac{bttda} iteratively extracts multiple blocks of
discriminative information, which can help in capturing more complex patterns
in neural data.
While \ac{bttda} is not specifically tailored to
gaze-independent \ac{bci} decoding, specific tensorizations of neural data can
offer advantages in coping with several obstacles arising from
gaze-independence.
The proposed \ac{bttda} method will be discussed in \autoref{sec:bttda}.

@@

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low \ac{snr} of the \ac{erp}.
\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
\ac{erp} latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic, iterative
method named \ac{wcble}.
This approach has been published by~\textcite{VanDenKerchove2024} and is presented
and evaluated on simulated data in \autoref{sec:wcble}.
We have also collected datasets to evaluate this approach on real \ac{erp} data
in gaze-independent settings.

@@

The interface can be operated in different \ac{vsa} conditions, as illustrated in
\autoref{fig:gaze/vsa}.

@@

In the overt case, users gazed at the cued
target they were also mentally attending; in the covert case, users
gazed at the center of the screen while mentally attending to the cued target.
We introduced a \ac{vsa} condition that is understudied in the context of
gaze-independent \ac{bci} development: split \ac{vsa}.
In split \ac{vsa}, the user mentally focuses on one cued target while gazing at
another (the distractor).
This last option has been scarcely studied but completes
the options to dissociate gaze and visuospatial attention, allowing us to
investigate the effect of (the lack of) gaze control on \ac{bci} performance.
This dataset and the performance of our proposed \ac{wcble} decoder have also been
published by \textcite{VanDenKerchove2024} and will be presented in
\autoref{sec:covert-align}.
Performance is also evaluated on a publicly available dataset.

@@

The hypothesis was that gaze-independent decoders, such as \ac{wcble}~\cite{VanDenKerchove2024},
would improve performance in covert and free \ac{vsa} conditions compared to standard
decoders like \ac{tlda}.
We aimed to determine if calibration in an overt \ac{vsa} setting could also enhance
performance in free and covert conditions, leveraging residual eye motor control during
calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
The results of this study are presented in \autoref{sec:patients}.

>>>>

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer
to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
\Cref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus,
\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).},
other eye tremors
\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.},
gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.},
involuntary movements, impaired pursuit, \ldots), and even complete
ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only affects vision and coordination but also
their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  \centering
  \makebox[\textwidth][c]{%
  \input{chapters/gaze_independence_incidence.tex}
  }
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

@@

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or it is very limited,
    such as the ability to communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups, as they are not
    able to use eye tracking solutions.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

The most severely impaired of these patients constitute the \ac{lis}
patient group, which forms one of the main \ac{bci} interest groups.
In this work, we use the term \ac{lis} for a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Hence, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.

@@

The general goal of this PhD is to tackle the gaze-dependence problem in the
visual oddball \ac{bci}.
\Cref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from gaze position, which do not necessarily need to coincide.

@@

To address this, we improved upon an in-house developed, state-of-the-art \ac{erp}
decoder, the \ac{stbf}~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(\acs{stbf-struct}).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, yielding insights for our next
efforts to cope with the non-stationarity of the P3 signal component.
This approach has been published by~\textcite{VanDenKerchove2022} and will be
described in \cref{sec:stbf-struct}.

@@

While \ac{hoda} is effective, it is limited by its reliance on a rigid core
structure, which may not fully capture the complexity of neural data in some
scenarios. To address this, \ac{bttda}
introduces a more flexible block-term structure.
\ac{bttda} iteratively extracts multiple blocks of
discriminative information, which can help in capturing more complex patterns
in neural data.
While \ac{bttda} is not specifically tailored to
gaze-independent \ac{bci} decoding, specific tensorizations of neural data can
offer advantages in coping with several obstacles arising from
gaze-independence.
The proposed \ac{bttda} method will be discussed in \cref{sec:bttda}.

@@

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low \ac{snr} of the \ac{erp}.
\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
\ac{erp} latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic, iterative
method named \ac{wcble}.
This approach has been published by~\textcite{VanDenKerchove2024} and is presented
and evaluated on simulated data in \cref{sec:wcble}.
We have also collected datasets to evaluate this approach on real \ac{erp} data
in gaze-independent settings.

@@

The interface can be operated in different \ac{vsa} conditions, as illustrated in
\cref{fig:gaze/vsa}.

@@

In the overt case, users gazed at the cued
target they were also mentally attending; in the covert case, users
gazed at the center of the screen while mentally attending to the cued target.
We introduced a \ac{vsa} condition that is understudied in the context of
gaze-independent \ac{bci} development: split \ac{vsa}.
In split \ac{vsa}, the user mentally focuses on one cued target while gazing at
another (the distractor).
This last option has been scarcely studied but completes
the options to dissociate gaze and visuospatial attention, allowing us to
investigate the effect of (the lack of) gaze control on \ac{bci} performance.
This dataset and the performance of our proposed \ac{wcble} decoder have also been
published by \textcite{VanDenKerchove2024} and will be presented in
\cref{sec:covert-align}.
Performance is also evaluated on a publicly available dataset.

@@

The hypothesis was that gaze-independent decoders, such as \ac{wcble}~\cite{VanDenKerchove2024},
would improve performance in covert and free \ac{vsa} conditions compared to standard
decoders like \ac{tlda}.
We aimed to determine if calibration in an overt \ac{vsa} setting could also enhance
performance in free and covert conditions, leveraging residual eye motor control during
calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
The results of this study are presented in \cref{sec:patients}.
